The Practical Application of Computerised Multistage Tests

### 2021-02-26 article.Rmd
Ability to estimate cut-score from MIFs

---
title: "The Practical Application of Computerised Multistage Tests"
author:
- name: James Page
  affiliation: SP Educational Consulting
  affiliation_url: https://www.speduconsulting.co.uk/
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    highlight: pygments
    latex_engine: lualatex
    fig_width: 5
version: 1
description: |
  A practical guide to the design and development of computerised multistage tests.
bibliography: references.bib
---

@@

@@

ggplot(plotData, aes(x=theta, y=trueScore)) + 
  geom_line(aes(colour=module)) +
  labs(title="Module Characteristic Curve", x="θ", y="True Score")
```
This is already showing us something a little weird. Our D (very easy) module is actually a little bit more difficult than our F (moderately hard) module at the top end of the scale. And the E (moderately easy) module is more difficult than the G (very hard) module. All of the modules are quite similar in terms of difficulty (the MCCs are quite close to each other), and moreover the lines are quite flat - there is little discrimination between them. We'll see how this can be expressed more obivously in the next section.

@@

In this section we'll show how to use the Module Information Function (MIF) to derive cut scores for module progression, and also how to build a scale scoring system for the test.

@@

The IIF can be represented visually like this, using the `plot` function from the `ltm` package:

@@

@@

Routing between modules is determined by the information functions for the modules in the next stage. We should aim to place the cut score to be that at which the MIFs cross.

```{r Stage 1 MIF}
plotData <- informationCoefficients(plot(raschmodel, 
                                         type = "IIC", 
                                         items = moduleA1_items, 
                                         plot = FALSE), 
                                    "A")

ggplot(plotData, aes(x=theta, y=information)) + 
  geom_line(aes(colour=module)) +
  labs(title="Module Information Function", x="θ", y="I(θ)")
```
You can see that the maximum information point Module A is pretty close to $\theta = 0$.

Now lets look at the MIFs for the two stage 2 modules:

```{r Stage 2 MIFs}
plotData <- rbind(
  informationCoefficients(plot(raschmodel, 
                               type = "IIC", 
                               items = moduleB1_items, plot = FALSE), 
                          "B"),
  informationCoefficients(plot(raschmodel, 
                               type = "IIC", 
                               items = moduleC1_items, plot = FALSE), 
                          "C"))

ggplot(plotData, aes(x=theta, y=information)) + 
  geom_line(aes(colour=module)) +
  labs(title="Module Information Function", x="θ", y="I(θ)") +
  geom_vline(xintercept = 0.81, linetype="dotted")
```
Here we can see that there is point of intersection between modules B and C. This suggests, that candidates who have a $\hat\theta < 0.81$ in Module A should be routed to Module B, and all others routed to Module C.

>>>>

---
title: "The Practical Application of Computerised Multistage Tests"
author:
- name: James Page
  affiliation: SP Educational Consulting
  affiliation_url: https://www.speduconsulting.co.uk/
date: "`r Sys.Date()`"
output:
  pdf_document:
    highlight: pygments
    latex_engine: lualatex
version: 1
description:
  A practical guide to the design and development of computerised multistage tests.
bibliography: references.bib
---

@@

### Rasch Model

We're now ready to fit the Rasch model, using the `rasch` function in the `ltm` package:

@@

ggplot(plotData, aes(x=theta, y=trueScore)) + 
  geom_line(aes(colour=module)) +
  labs(title="Module Characteristic Curve", x="θ", y="True Score")
```
This is already showing us something a little weird. Our D (very easy) module is actually a little bit more difficult than our F (moderately hard) module at the top end of the scale. And the E (moderately easy) module is more difficult than the G (very hard) module. All of the modules are quite similar in terms of difficulty (the MCCs are quite close to each other), and moreover the lines are quite flat - there is little discrimination between them. We'll see how this can be expressed more obviously in the next section.

@@

In this section we'll show how to use the Module Information Function (MIF) to derive cut scores for module progression, and also how to build a scale scoring system for the test. First we'll cover the topic of deriving scale scores for each module and then for the test.

### Scale Scores for Test Reporting

We could provide candidates with their $\hat{\theta}$ at test end but most candidates and testing organisations do not like theta values and the concept of zero as the population mean is difficult for most people. We therefore need to convert the theta value into a scale score. This usually involves fitting the theta to a range that is acceptable to the testing body (for example a score in the range 20-80).

Before we get to that point however we'll cover the case where the test administration system is unable to calculate the theta estimate and we need to provide a true score to scale score mapping. We do this by first converting true score to theta estimate, and then mapping the theta estimate on to a scale score. True score is simply the number correct for a given module.

### Converting True Score to Theta

Recall that the MCCs provide a mapping from the number of correct scores for each module to $\hat{\theta}$. A quintic polynomial can be used to fit the MCC, and the resulting coefficients then used by the test administration system to provide an ability estimate at the end of each module. The scale scores can then be used to provide an aggregated test score once the final stage of the MST has completed.

### Module A Mapping Function, $g_{A1}()$

First we use a polynomial regresssion to fit a set of coefficients to the Rasch model.

```{r Module A True Scores, echo=FALSE}
df_moduleA_TS <- trueScoreSummary(plot(raschmodel, type = "ICC", items = moduleA1_items, plot = FALSE))

# use a quintic polynomial regression model to make predictions
lm_moduleA_scores_poly <- lm(theta ~ poly(trueScore, degree=5, raw=TRUE), data = df_moduleA_TS)

# setup the score scale for module A
moduleA_TSRange <- data.frame(
  trueScore = c(seq(0,20)),
  theta = c(rep(0, 21))
)

# plot the theta/true score curve from the Rasch model
plot(x=df_moduleA_TS$theta, y=df_moduleA_TS$trueScore, 
     type="l", col = "coral1", main="Module A", xlab="θ", ylab="True Score")

# overlay the polynomial mapped true score points
scaleScores_moduleA_poly <- predict(lm_moduleA_scores_poly, moduleA_TSRange)
moduleA_TSRange$theta <- unname(scaleScores_moduleA_poly)
points(x=moduleA_TSRange$theta, y=moduleA_TSRange$trueScore, col="cornflowerblue", pch=4)

legend(2, 9, legend=c("MCC", "Quintic Poly Function"),
  col=c("coral1", "cornflowerblue"), lty=1:2, cex=0.8)

summary(lm_moduleA_scores_poly)
```

The model looks good for our use. From this we can get the coefficients (terms) that will be used for the polynomial function.

```{r Module A True Score Function}
options(scipen = 1, digits = 3)
gA_coeffs <- unname(coef(lm_moduleA_scores_poly))

(gA_coeffs)
```

The Avellain developers, when they want to calculate the ability for a candidate could use such a function to map from the number correct (NC) score to an estimated theta for candidate $i$: $\hat\theta_i$. For example, using R:

```{r}
gA1 <- function(x) {
  return(gA_coeffs[6]*x^5 + gA_coeffs[5]*x^4 + gA_coeffs[4]*x^3 + gA_coeffs[3]*x^2 + gA_coeffs[2]*x + gA_coeffs[1])
}
```

So, a candidate who scores 13 out of 20 on module A would have a $\hat\theta$ = `r gA1(13)`.

### Validating the Mapping Function
We can now use this function to overlay the original MCC:

```{r}
moduleA_TSRange <- data.frame(
  trueScore = c(seq(0,20)),
  theta = c(rep(0, 21))
)

# plot the theta/true score curve from the Rasch model
plot(x=df_moduleA_TS$theta, y=df_moduleA_TS$trueScore, 
     type="l", col = "coral1", main="Module A", xlab="θ", ylab="True Score")
moduleA_TSRange$theta <- gA1(moduleA_TSRange$trueScore)
points(x=moduleA_TSRange$theta, y=moduleA_TSRange$trueScore, col="purple", pch=4)
legend(2, 9, legend=c("MCC", "gA1()"),
  col=c("coral1", "purple"), lty=1:2, cex=0.8)
```
We can see that it is a good fit, both in the middle of the range, and importantly at the upper and lower bounds.

### Module D-H Mapping Functions, $g_{D1}(), g_{E1}(), g_{F1}, g_{G1}(), g_{H1}()$

```{r}
df_moduleD_TS <- trueScoreSummary(plot(raschmodel, type = "ICC", items = moduleD1_items, plot = FALSE))
lm_moduleD_scores_poly <- lm(theta ~ poly(trueScore, degree=5, raw=TRUE), data = df_moduleD_TS)
gD_coeffs <- unname(coef(lm_moduleD_scores_poly))

df_moduleE_TS <- trueScoreSummary(plot(raschmodel, type = "ICC", items = moduleE1_items, plot = FALSE))
lm_moduleE_scores_poly <- lm(theta ~ poly(trueScore, degree=5, raw=TRUE), data = df_moduleE_TS)
gE_coeffs <- unname(coef(lm_moduleE_scores_poly))

df_moduleF_TS <- trueScoreSummary(plot(raschmodel, type = "ICC", items = moduleF1_items, plot = FALSE))
lm_moduleF_scores_poly <- lm(theta ~ poly(trueScore, degree=5, raw=TRUE), data = df_moduleF_TS)
gF_coeffs <- unname(coef(lm_moduleF_scores_poly))

df_moduleG_TS <- trueScoreSummary(plot(raschmodel, type = "ICC", items = moduleG1_items, plot = FALSE))
lm_moduleG_scores_poly <- lm(theta ~ poly(trueScore, degree=5, raw=TRUE), data = df_moduleG_TS)
gG_coeffs <- unname(coef(lm_moduleG_scores_poly))

df_moduleH_TS <- trueScoreSummary(plot(raschmodel, type = "ICC", items = moduleH1_items, plot = FALSE))
lm_moduleH_scores_poly <- lm(theta ~ poly(trueScore, degree=5, raw=TRUE), data = df_moduleH_TS)
gH_coeffs <- unname(coef(lm_moduleH_scores_poly))
```

```{r Output g Functions}
panel1_coeffs <- data.frame(t(gA_coeffs))
panel1_coeffs <- rbind(panel1_coeffs, data.frame(t(gD_coeffs)))
panel1_coeffs <- rbind(panel1_coeffs, data.frame(t(gE_coeffs)))
panel1_coeffs <- rbind(panel1_coeffs, data.frame(t(gF_coeffs)))
panel1_coeffs <- rbind(panel1_coeffs, data.frame(t(gG_coeffs)))
panel1_coeffs <- rbind(panel1_coeffs, data.frame(t(gH_coeffs)))
panel1_coeffs$moduleID <- c('A1', 'D1', 'E1', 'F1', 'G1', 'H1')
panel1_coeffs <- panel1_coeffs[, c(7, 6, 5, 4, 3, 2, 1)]

# OUTPUT a CSV file too...
write_csv(panel1_coeffs, "../data/scoring-functions-panel1.csv")
```

We use the concept of the information function as a way of determining the best cut scores to use. Where a choice is needed to route onward to the next stage, we need to select a cut score to use. We could do this by simply selecting a mid-point in the true scale. For example, if the routing module A has 12 items, then we might assume that if a candidate scores 0-5 they route to the easy module in the next stage (module B), and if they score 6-12 they route to the harder module (module C). Ideally though we'd like to base this on the results of the Rasch analysis. Using the Item Information Function (IIF) gives us a way of selecting a cut score based on IRT. First lets explain the Item Informtation Function.

@@

We mentioned earlier that the IIF provides a way of showing at what ability level an item is most discriminating. The IIF can be represented visually like this, using the `plot` function from the `ltm` package:

@@

For example this item shows a peak information for $\hat{\theta} =$ `r itemPlot$theta[which.max(itemPlot$information)]`.

@@

Routing between modules can then determined by cut scores derived from the intersection of information functions for the modules in the next stage. We should aim to place the cut score to be that at which the MIFs cross.

Let's look at the MIFs for the two stage 2 modules:

```{r Stage 2 MIFs}
iifB <- informationCoefficients(plot(raschmodel, type = "IIC", 
                               items = moduleB1_items, plot = FALSE), 
                               "B")
iifC <- informationCoefficients(plot(raschmodel, type = "IIC", 
                               items = moduleC1_items, plot = FALSE), 
                               "C")

plotData <- rbind(iifB, iifC)

modIIF <- inner_join(iifB, iifC, by = "theta") %>%
  rename(moduleB = information.x, moduleC = information.y) %>%
  dplyr::select(theta, moduleB, moduleC) %>%
  arrange((abs(moduleB - moduleC))) %>%
  head(n = 1)
cutTheta <- modIIF$theta

ggplot(plotData, aes(x=theta, y=information)) + 
  geom_line(aes(colour=module)) +
  labs(title="Module Information Function", x="θ", y="I(θ)") +
  geom_vline(xintercept = cutTheta, linetype="dotted")
```
Here we can see that there is point of intersection between modules B and C where $\hat\theta = $`r cutTheta`. If we read off the true score value that is equivalent for this theta estimate in Module A, then we can use this as the cut score. Below this value the candidate should be routed to Module B, above it they should be routed to Module C.
